Does at-issueness predict projection? Further investigations of the Gradient Projection Principle

### 2022-10-26 writing/investigating-GPP.tex
updated paper section 2

@@

hallo

Exps.~1-3 were designed to investigate whether not-at-issueness and projection are positively correlated, as predicted by the Gradient Projection Principle, as well as to compare various measures of at-issueness.\footnote{\label{f-github}The experiments, data, and R code for generating the figures and analyses of the experiments reported in this paper are available at \url{https://github.com/judith-tonhauser/projective-probability}. All experiments were conducted with IRB approval. Exps.~1b, 2b, and 3b were preregistered: \url{https://osf.io/cxq47}.}

To measure projection, Exps.~1-3 employed the `certain that' diagnostic for projection, which is suitable to measure projection for both polar questions and declarative sentences (see also, e.g.\ \citealt{tonhauser-salt26,djaerv-bacovcin-salt27,stevens-etal2017,lorson2018,tbd-variability,mahler-nels,mahler2020,demarneffe-etal-sub23}).\footnote{For other diagnostics for projection see, e.g.\ \citealt{smith-hall11,xue-onea11}, and \citealt{brst-lang11}; see also the discussion in \citealt{tbd-variability}.}  On this diagnostic, participants are presented with an utterance, like one of the utterances in (\ref{proj-stim}), in which the content to be diagnosed for projection (here, that Julian dances salsa) occurs embedded under an entailment-canceling operator:

\begin{exe}

\ex\label{proj-stim} `certain that' diagnostic for projection
\begin{xlist}
\ex Daniel: Did Cole discover that Julian dances salsa?
\ex Daniel: Cole didn't discover that Julian dances salsa.
\ex Daniel: Perhaps Cole discovered that Julian dances salsa.
\ex Daniel: If Cole discovered that Julian dances salsa, Logan will be joyful.
\end{xlist}
\end{exe}
To assess projection, participants are asked whether the speaker is certain of the content; for instance, in \ref{proj-stim}, participants are asked whether Daniel is certain that Julian dances salsa. If a participant takes the speaker to be certain of the content, we assume that the content projects; if a participant does not take the speaker to be certain of the content, we assume that the content does not project. Following \citealt{tbd-variability}, the `certain that' diagnostic was implemented in Exps.~1-3 with a gradient response scale: participants gave their responses on a slider marked `no' at one end and `yes' at the other. We assume that the closer to `yes' a participant's response is, the more the speaker is certain of the content, that is, the more projective the content is. In other words, we assume that gradient certainty ratings reflect gradience in the speaker's commitment to the truth of the content.\footnote{Strictly speaking,  gradient certainty ratings reflect gradience in the degree to which participants \emph{perceive} the speaker to be committed to the truth of the content. That is, as in any experiment, the quantity of interest, in this case speaker commitment, is only indirectly measured.} As discussed in \citealt{tbd-variability}, a second interpretation of gradient certainty ratings is that they reflect a participant's degree of belief in the speaker's (binary) commitment to the truth of the content. While we adopt the first interpretation in our discussion, we remain agnostic about the interpretation of gradient certainty ratings.

At-issueness was measured with different diagnostics across Exps.~1-3. For polar questions (Exps.~1q, 2q, 3q), at-issueness was assessed with the three measures in (\ref{ai-question}):

\begin{exe}
\ex\label{ai-question} At-issueness diagnostics for polar questions
\begin{xlist}
\ex `asking whether $c$' 
\\ {\bf Christopher:} Did Cole discover that Julian dances salsa?
\\ Is Christopher asking whether Julian dances salsa?

\ex {\em yes}, $c$
\\ {\bf Christopher:} Did Cole discover that Julian dances salsa?
\\ {\bf Sandy:} Yes, Julian dances salsa.
\\ Does Sandy's response to Christoper sound good?

\ex {\em yes, but} $\neg c'$
\\ {\bf Christopher:} Did Cole discover that Julian dances salsa?
\\ {\bf Sandy:} Yes, but Cole didn't discover it. 
\\ Does Sandy's response to Christoper sound good?

@@

For declaratives, at-issueness was assessed with the three measures in (\ref{ai-decl}):

\begin{exe}
\ex\label{ai-decl} At-issueness diagnostics for declaratives

\ex `sure that $c$'
\\ {\bf Christopher:} Cole didn't discover that Julian dances salsa.
\\ \hspace*{2cm} Perhaps Cole discovered that Julian dances salsa.
\\ {\bf Sandy:} Are you sure?
\\ {\bf Christopher:} Yes, I'm sure that Julian dances salsa.
\\ Did Christopher answer Sandy's question?

\ex {\em yes, that's true}, $c$
\\ {\bf Christopher:} Cole didn't discover that Julian dances salsa.
\\ {\bf Sandy:} Yes, that's true, Julian dances salsa.
\\ Does Sandy's response to Christopher sound good?

\ex {\em yes, that's true, but} $\neg c'$
\\ {\bf Christopher:} Cole didn't discover that Julian dances salsa.
\\ {\bf Sandy:} Yes, that's true, but Cole discovered it. 
\\ Does Sandy's response to Christopher sound good?

@@

{\bf Exp.} & {\bf Embedding} & {\bf Projection measure} & {\bf At-issueness measure} \\ 
\hline
1q & question & certain that $c$ & asking whether $c$ \\
1n & negation & certain that $c$ & sure that $c$ \\
1m & modal & certain that $c$ & sure that $c$ \\
1c & conditional & certain that $c$ & sure that $c$ \\
\hline
2q & question & certain that $c$ & {\em yes}, $c$ \\ 
2n & negation & certain that $c$& {\em yes, that's true}, $c$ \\ 
2m & modal & certain that $c$& {\em yes, that's true}, $c$ \\ 
2c & conditional & certain that $c$& {\em yes, that's true}, $c$ \\ 
\hline
3q & question & certain that $c$& {\em yes}, but $\neg c'$ \\ 
3n & negation & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
3m & modal & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
3c & conditional & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
\hline
\end{tabular}
\caption{Overview of the entailment-canceling embeddings, projection measures, and at-issueness measures in Exps.~1-3, where $c$ is the content under investigation (in the target trials: the content of the clausal complement) and $c'$ is a different, salient content (in the target trials: the main clause content).}\label{t-overview}
\end{table}

{\bf Comparisons for at-issueness:}

\item Across Exps 1q, 2q, 3q (same assumption about at-issue content partitioning context set; reveals effect of embedding)

\item Across Exps n, m, c (same measure of at-issueness in Exps 1, 2, 3; reveals effect of embedding)

\item 2q -- 2n, 2m, 2c (yes...$c$; i.e., anaphor with $c$)

\item 3q -- 3n, 3m, 3c (yes...but $\neg c$, i.e., anaphor with $c$)

\footnote{Exps.~1-3 also allow us to address the question of whether the projection of contents embedded under different entailment-canceling operators is positively correlated, as assumed in the semantics/pragmatics literature (e.g., REFERENCES). Prior experimental literature: \citealt{smith-hall11,smith-hall-cls,demarneffe-etal-sub23}. As shown in Supplement \ref{a-projection}, this is the case.}

@@

Exps.~1-3 investigated the projection and at-issueness of the CC of the 20 clause-embedding predicates in (\ref{pred}). The clausal complements of the 20 predicates were realized by 20 clauses (provided in Supplement \ref{a-clauses}), for a total of 400 predicate/clause combinations. These 400 predicate/clause combinations were combined with proper name subjects (a unique name per complement clause) to realize polar questions (in Exps. 1q, 2q, and 3q), negated declaratives (in Exps. 1n, 2n, and 3n), declaratives with the modal adverb {\em perhaps} (in Exps. 1m, 2m, and 3m), and the antecedents of conditionals (in Exps. 1c, 2c, and 3c). These sentences were presented as utterances by a named speaker. A sample set of target utterances is given in (\ref{sample-stims}): here, the named speaker is {\em Daniel}, the subject of the clause-embedding predicate is {\em Cole}, the predicate is {\em discover}, and the complement clause is {\em Julian dances salsa}.\footnote{The indirect object of {\em inform} was realized by the proper name {\em Sam}. The predicates differed in the tense in which they were realized. In polar interrogatives as well as in negated and modalized declaratives, eventive predicates, like {\em discover} and {\em hear}, were realized in the past tense and stative predicates, like {\em know} and {\em be annoyed}, were realized in the present tense.  In the antecedents of conditionals, all predicates were realized in the present tense. For the consequents of the conditional sentences see Supplement \ref{a-target}.}

\begin{exe}
\ex\label{sample-stims}
\begin{xlist}
\ex {\bf Daniel:} ``{\em Did Cole discover that Julian dances salsa?}"
\ex {\bf Daniel:} ``{\em Cole didn't discover that Julian dances salsa.}"
\ex {\bf Daniel:} ``{\em Perhaps Cole discovered that Julian dances salsa.}"
\ex {\bf Daniel:} ``{\em If Cole discovered that Julian dances salsa, Logan will be joyful.}"
\end{xlist}
\end{exe}
The proper names that realized the speakers, the subjects of the clause-embedding predicates, the subjects of the complement clauses, and the subjects of the consequents  were all unique.

Projection and at-issueness of the contents of the clausal complements were measured in separate blocks. In the projection blocks across Exps.~1-3, projection was measured with the `certain that' diagnostic. The target stimuli consisted of the target utterances, as shown in Figure \ref{f-projection-trials}.

@@

\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=4.4cm,width=7.6cm]{figures/2q-proj}}
\caption{Exps.~1q, 2q, and 3q.}\label{fig-exp1q-projection}
\end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\centering
\fbox{\includegraphics[height=4.4cm,width=7.6cm]{figures/1n-proj}} 
\caption{Exps.~1n, 2n, and 3n.}\label{fig-exp1n-projection}
 \end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=4.4cm,width=7.6cm]{figures/1m-proj}}
\caption{Exps.~1m, 2m, and 3m}\label{fig-exp1m-projection}
 \end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\centering
\fbox{\includegraphics[height=4.4cm,width=7.6cm]{figures/1c-proj}} 
\caption{Exps.~1c, 2c, and 3c}\label{fig-exp1c-projection}
\end{subfigure}

\caption{Target trials in projection blocks of Exps.~1, 2, and 3 for the complement {\em Julian dances salsa}.}\label{f-projection-trials}
\end{figure}

@@

\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/1q-ai}}
\caption{Exp.~1q.}\label{fig-exp1q-ai}
\end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
\centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/2q-ai}} 
\caption{Exp.~2q.}\label{fig-exp2q-ai}
 \end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/3q-ai}}
\caption{Exp.~3q}\label{fig-exp3q-ai}
 \end{subfigure}

\caption{Target trials in at-issueness blocks of Exps.~1q, 2q, and 3q for the complement {\em Julian dances salsa}.}\label{f-ai-trialsq}
\end{figure}

@@

\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/2n-ai}}
\caption{Exp.~2n.}\label{fig-exp1q-ai}
\end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/2m-ai}} 
\caption{Exp.~2m.}\label{fig-exp2q-ai}
 \end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=9cm]{figures/2c-ai}}
\caption{Exp.~2c}\label{fig-exp3q-ai}
 \end{subfigure}

\caption{Target trials in at-issueness blocks of Exps.~2n, 2m, and 2c for the complement {\em Julian dances salsa}.}\label{f-ai-trialsq}
\end{figure}

@@

\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/3n-ai}}
\caption{Exp.~3n.}\label{fig-exp1q-ai}
\end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/3m-ai}} 
\caption{Exp.~3m.}\label{fig-exp2q-ai}
 \end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=9cm]{figures/3c-ai}}
\caption{Exp.~3c}\label{fig-exp3q-ai}
 \end{subfigure}

\caption{Target trials in at-issueness blocks of Exps.~3n, 3m, and 3c for the complement {\em Julian dances salsa}.}\label{f-ai-trialsq}
\end{figure}

In the projection block, target stimuli consisted of a fact and a polar question that was ut- tered by a named speaker, as shown in Figure 1B. The polar questions were formed by real- izing the 20 clauses as the complements of the 20 clause-embedding predicates in Figure 1C. Participants were told to imagine that they are at a party and that, on walking into the kitchen, they overhear somebody ask somebody else a question. Projection was measured using the ?certain that? diagnostic (Dj\"arv \& Bacovcin, 2017; Lorson, 2018; Mahler, 2020; Tonhauser et al., 2018): participants were asked to rate whether the speaker was certain of the CC, taking into consideration the fact. They gave their responses on a slider marked ?no? at one end (coded as 0) and ?yes? at the other (coded as 1). Greater speaker commitment to the CC should result in higher slider ratings.

Participants were told to imagine that they are at a party and that, on walking into the kitchen, they overhear somebody ask a question. Participants were asked to rate whether the speaker was certain of the CC. They gave their responses on a slider marked `no' at one end (coded as 0) and `yes' at the other (coded as 1), as shown in Figure \ref{fig-trial-exp1}.

After completing the experiment, participants filled out a short, optional survey about their age, their gender, their native language(s) and, if English is their native language, whether they are a speaker of American English (as opposed to, e.g.\ Australian or Indian English). To encourage truthful responses, participants were told that they would be paid no matter what answers they gave in the survey.

@@

\section{Data exclusion}\label{a-participants}

This supplement provides information on the recruited participants, the criteria by which participants' data were excluded, the remaining participants (that is, the participants whose data entered into the analysis), and the payment for each of Exps.~1-3. Table \ref{f-exclusion} provides information on the recruited and remaining participants, including the total number, the range of the participants' ages, the mean ages of the participants, and their self-reported gender (`f' = female, `m' = male, `o' = other, `u' = undeclared). No gender data was collected in Exp.~1q. Participants' data were excluded based on the following criteria:

@@

\item `multiple': Due to an experimental glitch, some participants participated more than once. Since no information was available on which one was their first take, those participants' data was removed (which means that for each participant who took the experiment twice, at least two data sets were removed).

@@

\item `controls': Participants' data were excluded if their mean rating on the 6 main clause control items in the projection block was more than 2 sd above the group mean (in Exps.~1q, 2q, and 3q) or more than 2 sd below the group mean (in the remaining experiments). Participants' data were also excluded if their mean rating on the 6 main clause control items in the at-issueness block was more than 2 sd above the group mean (across Exps.~1-3).

>>>>

projection of contents embedded under different entailment-canceling operators is positively correlated, as assumed in the semantics/pragmatics literature (e.g., REFERENCES). Prior experimental literature: \citealt{smith-hall11,smith-hall-cls,demarneffe-etal-sub23}.

@@

Exps.~1-3 were designed to investigate whether not-at-issueness and projection are positively correlated, as predicted by the Gradient Projection Principle, to compare the projection of contents from different entailment-canceling environments, and to compare different measures of at-issueness.\footnote{\label{f-github}The experiments, data, and R code for generating the figures and analyses of the experiments reported in this paper are available at \url{https://github.com/judith-tonhauser/projective-probability}. All experiments were conducted with IRB approval. Exps.~1b, 2b, and 3b were preregistered: \url{https://osf.io/cxq47}.} To be able to compare across the experiments, the experiments investigated the same contents and each of the experiments consisted of two blocks: a projection block, in which the projection of the contents was measured, and an at-issueness block, in which the at-issueness of the contents was measured. The at-issueness blocks of Exps.~1-3 implemented different measures of at-issueness, which allows for a comparison of at-issueness measures. To compare projection from different entailment-canceling environments, each of Exps.~1-3 consisted of four experiments (for a total of 12 experiments) where the contents were realized by expressions in a polar question, in the scope of sentential negation, in the scope of the epistemic possibility modal {\em perhaps}, or in the antecedent of a conditional. For an overview of the 12 experiments see Table \ref{t:overview}: the name of each of the 12 experiments identifies whether belongs to the set of Exps.~1, 2, or 3 (which differ in at-issueness measure) and whether the embedding-environment is a polar question (`q'), negation (`n'), the epistemic possibility modal (`m'), or the antecedent of a conditional (`c').

Projection was measured across all 12 experiments with the `certain that' diagnostic, which has been used to measure projection with both polar interrogative and declarative sentences (see, e.g., \citealt{tonhauser-salt26,djaerv-bacovcin-salt27,stevens-etal2017,lorson2018,tbd-variability,mahler-nels,mahler2020,demarneffe-etal-sub23}).\footnote{For other diagnostics for projection see, e.g.\ \citealt{smith-hall11,xue-onea11}, and \citealt{brst-lang11}; see also the discussion in \citealt{tbd-variability}.}  On this diagnostic, participants are presented with an utterance, like one of Daniel's utterances in (\ref{proj-stim}), in which the content to be diagnosed for projection (here, that Julian dances salsa) occurs in one of the four entailment-canceling environments:

\begin{exe}
\ex\label{proj-stim} `certain that' diagnostic for projection
\begin{xlist}
\ex Daniel: Did Cole discover that Julian dances salsa? \hfill [Exps.~1q, 2q, 3q]
\ex Daniel: Cole didn't discover that Julian dances salsa. \hfill [Exps.~1n, 2n, 3n]
\ex Daniel: Perhaps Cole discovered that Julian dances salsa. \hfill [Exps.~1m, 2m, 3m]
\ex Daniel: If Cole discovered that Julian dances salsa, Logan will be joyful. \hfill [Exps.~1c, 2c, 3c]
\end{xlist}
Question to participants: Is Daniel certain that Julian dances salsa?
\end{exe}
To measure projection, participants are asked whether the speaker is certain of the relevant content. In \ref{proj-stim}, for instance, participants are asked whether Daniel is certain that Julian dances salsa. If a participant takes the speaker to be certain of the content, we assume that the content projects out of the entailment-canceling environment; if a participant does not take the speaker to be certain of the content, we assume that the content does not project.

At-issueness was measured with different diagnostics across Exps.~1-3. In Exps.~1, at-issueness was measured with the `asking whether' diagnostic in Exp.~1q and with the `sure that' diagnostic in Exps.~1n, 1m, and 1c.

\begin{exe}
\ex\label{ai-exps1} At-issueness measures in Exps.~1
\begin{xlist}
\ex `asking whether' \\ Christopher: Did Cole discover that Julian dances salsa? \hfill [Exp.~1q]
\\ Question to participants: Is Christopher asking whether Julian dances salsa?

\ex `sure that' \\ Christopher: Cole didn't discover that Julian dances salsa.  \hfill [Exp.~1n]
\\ \hspace*{2cm} Perhaps Cole discovered that Julian dances salsa.  \hfill [Exp.~1m]
\\ \hspace*{2cm} If Cole discovered that Julian dances salsa, Logan will be joyful.  \hfill [Exp.~1c]
\\ Sandy: Are you sure?
\\ Christopher: Yes, I'm sure that Julian dances salsa.
\\ Question to participants: Did Christopher answer Sandy's question?

@@

In Exps.~2, at-issueness was measured with an assent diagnostic, specifically, with the version of the assent diagnostic in which the assent is followed with the content to be diagnosed. Participants are presented with a dialogue in which the first speaker's utterance gives rise to the content to be diagnosed (here, that Julian dances salsa). The second speaker responds in the affirmative (for polar interrogatives, as shown in (\ref{ai-exps2}a)) or assents (for declaratives, as shown in (\ref{ai-exps2}b). The affirmation/assent is followed by the content to be diagnosed, here that Julian dances salsa.

\begin{exe}
\ex\label{ai-exps2} At-issueness measures in Exps.~2
\begin{xlist}
\ex {\em yes}, $c$
\\ Christopher: Did Cole discover that Julian dances salsa? \hfill [Exp.~2q]
\\ Sandy: Yes, Julian dances salsa.
\\ Question to participants: Does Sandy's response to Christoper sound good?

\ex {\em yes, that's true}, $c$
\\ Christopher: Cole didn't discover that Julian dances salsa.  \hfill [Exp.~2n]
\\ \hspace*{2cm} Perhaps Cole discovered that Julian dances salsa.  \hfill [Exp.~2m]
\\ \hspace*{2cm} If Cole discovered that Julian dances salsa, Logan will be joyful.  \hfill [Exp.~2c]
\\ Sandy: Yes, that's true, Julian dances salsa.
\\ Question to participants: Does Sandy's response to Christopher sound good?

\end{xlist}
\end{exe}
Participants are asked whether the second speaker's response is acceptable in response to the first speaker's utterance. If a participant judges the response to be acceptable, the relevant content is taken to be at-issue. If a participant judges the response to be unacceptable, the relevant content is taken to be not at-issue.

In Exps.~3, at-issueness was measured with another assent diagnostic, specifically, with the version of the assent diagnostic in which the assent is followed with the denial of another, salient content. As in the other assent diagnostic, participants are presented with a dialogue in which the first speaker's utterance gives rise to the content to be diagnosed (here, that Julian dances salsa). The second speaker responds in the affirmative (for polar interrogatives, as shown in (\ref{ai-exps3}a)) or assents (for declaratives, as shown in (\ref{ai-exps3}b). The affirmation/assent is followed by the denial of another salient content, here, that Cole (didn't) discover it.

\begin{exe}
\ex\label{ai-exps3} At-issueness measures in Exps.~3
\begin{xlist}
\ex {\em yes}, $c$
\\ Christopher: Did Cole discover that Julian dances salsa? \hfill [Exp.~3q]
\\ Sandy: Yes, but Cole didn't discover it. {\bf or that?}
\\ Question to participants: Does Sandy's response to Christoper sound good?

\ex {\em yes, that's true}, $c$
\\ Christopher: Cole didn't discover that Julian dances salsa.  \hfill [Exp.~3n]
\\ \hspace*{2cm} Perhaps Cole discovered that Julian dances salsa.  \hfill [Exp.~3m]
\\ \hspace*{2cm} If Cole discovered that Julian dances salsa, Logan will be joyful.  \hfill [Exp.~3c]
\\ Sandy: Yes, that's true, but Cole discovered / didn't discover it. {\bf or that?}
\\ Question to participants: Does Sandy's response to Christopher sound good?

\end{xlist}
\end{exe}
As with the other assent diagnostic, participants are asked whether the second speaker's response is acceptable in response to the first speaker's utterance. If a participant judges the response to be acceptable, the relevant content is taken to be at-issue. If a participant judges the response to be unacceptable, the relevant content is taken to be not at-issue.

@@

{\bf Exp.} & {\bf Embedding} & {\bf Projection measure} & {\bf At-issueness measure} \\ 
\hline
1q & question & certain that $c$ & asking whether $c$ \\
1n & negation & certain that $c$ & sure that $c$ \\
1m & modal & certain that $c$ & sure that $c$ \\
1c & conditional & certain that $c$ & sure that $c$ \\
\hline
2q & question & certain that $c$ & {\em yes}, $c$ \\ 
2n & negation & certain that $c$& {\em yes, that's true}, $c$ \\ 
2m & modal & certain that $c$& {\em yes, that's true}, $c$ \\ 
2c & conditional & certain that $c$& {\em yes, that's true}, $c$ \\ 
\hline
3q & question & certain that $c$& {\em yes}, but $\neg c'$ \\ 
3n & negation & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
3m & modal & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
3c & conditional & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
\hline
\end{tabular}
\caption{Overview of Exps.~1-3, which differ in the entailment-canceling embedding (q = polar question, n = negation, m = epistemic possibility modal, c = antecedent of conditional) and measures for not-at-issueness. $c$ is the content under investigation (in the target trials: the content of the clausal complement) and $c'$ is a different, salient content (in the target trials: the main clause content).}\label{t:overview}
\end{table}

@@

The projection and at-issueness of the contents of 20 clauses (such as {\em Julian dances salsa}) were measured in separate blocks in Exps.~1-3. For the full set of clauses, see Appendix \ref{a-clauses}.

In the projection blocks of Exps.~1-3, target stimuli consisted of a polar question sentence, a negated sentence, a sentence with {\em perhaps}, or a conditional sentence that was uttered by a named speaker, as shown in Fig.~\ref{fig:projection-trials}. Each sentence was formed by realizing the 20 clauses as the complements of the 20 clause-embedding predicates in Fig.~\ref{fig:predicates}.\footnote{The proper names that realized the speakers, the subjects of the clause-embedding predicates, the subjects of the complement clauses, and the subjects of the consequents  were all unique. The indirect object of {\em inform} was realized by the proper name {\em Sam}. The predicates differed in the tense in which they were realized. In polar interrogatives as well as in negated and modalized declaratives, eventive predicates, like {\em discover} and {\em hear}, were realized in the past tense and stative predicates, like {\em know} and {\em be annoyed}, were realized in the present tense.  In the antecedents of conditionals, all predicates were realized in the present tense. For the consequents of the conditional sentences see Supplement \ref{a-target}.} Participants were told to imagine that they are at a party and that, on walking into the kitchen, they overhear somebody ask somebody else a question. Projection was measured using the `certain that' diagnostic \cite{djaerv-bacovcin-salt27,tbd-variability, lorson2018, mahler2020}: participants were asked to rate whether the speaker was certain of the CC, taking into consideration the fact. They gave their responses on a slider marked `no' at one end (coded as 0) and `yes' at the other (coded as 1). Greater speaker commitment to the CC should result in higher slider ratings.

%Following \citealt{tbd-variability}, the `certain that' diagnostic was implemented in Exps.~1-3 with a gradient response scale: participants gave their responses on a slider marked `no' at one end and `yes' at the other. We assume that the closer to `yes' a participant's response is, the more the speaker is certain of the content, that is, the more projective the content is. In other words, we assume that gradient certainty ratings reflect gradience in the speaker's commitment to the truth of the content.\footnote{Strictly speaking,  gradient certainty ratings reflect gradience in the degree to which participants \emph{perceive} the speaker to be committed to the truth of the content. That is, as in any experiment, the quantity of interest, in this case speaker commitment, is only indirectly measured.} As discussed in \citealt{tbd-variability}, a second interpretation of gradient certainty ratings is that they reflect a participant's degree of belief in the speaker's (binary) commitment to the truth of the content. While we adopt the first interpretation in our discussion, we remain agnostic about the interpretation of gradient certainty ratings.

\begin{figure}[h!]
\centering
\fbox{\begin{minipage}{15cm}{\em acknowledge, admit, announce, be annoyed, be right, confess, confirm, demonstrate, discover, establish, hear, inform, know, pretend, prove, reveal, say, see, suggest, think. }\end{minipage}}
\caption{20 clause-embedding predicates in Exps.~1-3.}\label{fig:predicates}
\end{figure}

@@

\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=4.4cm,width=7.6cm]{figures/2q-proj}}
\caption{Exps.~1q, 2q, and 3q.}\label{fig-exp1q-projection}
\end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\centering
\fbox{\includegraphics[height=4.4cm,width=7.6cm]{figures/1n-proj}} 
\caption{Exps.~1n, 2n, and 3n.}\label{fig-exp1n-projection}
 \end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=4.4cm,width=7.6cm]{figures/1m-proj}}
\caption{Exps.~1m, 2m, and 3m}\label{fig-exp1m-projection}
 \end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\centering
\fbox{\includegraphics[height=4.4cm,width=7.6cm]{figures/1c-proj}} 
\caption{Exps.~1c, 2c, and 3c}\label{fig-exp1c-projection}
\end{subfigure}
\caption{Target trials in projection blocks of Exps.~1, 2, and 3 for the complement {\em Julian dances salsa}.}\label{fig:projection-trials}
\end{figure}

In the at-issueness blocks of Exps.~1, target

In the projection blocks of Exps.~1-3, target stimuli consisted of a polar question sentence, a negated sentence, a sentence with {\em perhaps}, or a conditional sentence that was uttered by a named speaker, as shown in Fig.~\ref{fig:projection-trials}. Each sentence was formed by realizing the 20 clauses as the complements of the 20 clause-embedding predicates in Fig.~\ref{fig:predicates}.\footnote{The proper names that realized the speakers, the subjects of the clause-embedding predicates, the subjects of the complement clauses, and the subjects of the consequents  were all unique. The indirect object of {\em inform} was realized by the proper name {\em Sam}. The predicates differed in the tense in which they were realized. In polar interrogatives as well as in negated and modalized declaratives, eventive predicates, like {\em discover} and {\em hear}, were realized in the past tense and stative predicates, like {\em know} and {\em be annoyed}, were realized in the present tense.  In the antecedents of conditionals, all predicates were realized in the present tense. For the consequents of the conditional sentences see Supplement \ref{a-target}.} Participants were told to imagine that they are at a party and that, on walking into the kitchen, they overhear somebody ask somebody else a question. Projection was measured using the `certain that' diagnostic \cite{djaerv-bacovcin-salt27,tbd-variability, lorson2018, mahler2020}: participants were asked to rate whether the speaker was certain of the CC, taking into consideration the fact. They gave their responses on a slider marked `no' at one end (coded as 0) and `yes' at the other (coded as 1). Greater speaker commitment to the CC should result in higher slider ratings.

@@

\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/1q-ai}}
\caption{Exp.~1q.}
\end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/1n-ai}} 
\caption{Exp.~1n.}
 \end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/1m-ai}}
\caption{Exp.~1m}
 \end{subfigure}%
 \begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/1c-ai}}
\caption{Exp.~1c}
 \end{subfigure}

\caption{Target trials in at-issueness blocks of Exps.~1 for the complement {\em Julian dances salsa}.}\label{fig:ai-exps1}
\end{figure}

@@

\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/2q-ai}}
\caption{Exp.~2q.}
\end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/2n-ai}} 
\caption{Exp.~2n.}
 \end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/2m-ai}}
\caption{Exp.~2m}
 \end{subfigure}%
 \begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/2c-ai}}
\caption{Exp.~2c}
 \end{subfigure}

\caption{Target trials in at-issueness blocks of Exps.~2 for the complement {\em Julian dances salsa}.}\label{fig:ai-exps3}
\end{figure}

@@

\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/3q-ai}}
\caption{Exp.~3q.}
\end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/3n-ai}} 
\caption{Exp.~3n.}
 \end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/3m-ai}}
\caption{Exp.~3m}
 \end{subfigure}%
 \begin{subfigure}[t]{0.5\textwidth}
        \centering
\fbox{\includegraphics[height=5.4cm,width=7.6cm]{figures/3c-ai}}
\caption{Exp.~3c}
 \end{subfigure}

\caption{Target trials in at-issueness blocks of Exps.~3 for the complement {\em Julian dances salsa}.}\label{fig:ai-exps3}
\end{figure}

====

The projection block also included 6 control trials, which functioned as attention checks. The content of these stimuli was expected not to project: For example, in Fig.~\ref{fig-exp1-projection-control}, the speaker is not committed to the main clause content, that Samantha has a new hat. The same 6 main clauses were also used to form 6 filler trials in the prior block; a sample stimulus is given in Fig.~\ref{fig-exp1-prior-filler}. These filler stimuli were not used to assess participants' attention. For the full set of stimuli see Supplement \ref{a-stim}.

Each participant's stimulus set was semi-randomly generated by first randomly pairing up the 20 predicates and clauses. Half of the stimuli were then randomly assigned the respective clause's higher-probability fact, and half its lower-probability fact. Participants completed a total of 52 trials: 20 target trials in each block, 6 control trials in the projection block, and 6 filler trials in the prior block. Each participant completed the same 6 filler and control trials. Block order and within-block trial order were randomized.

After completing the experiment, participants filled out a short optional demographic survey. To encourage truthful responses, participants were told that they would be paid no matter what answers they gave in the survey.

@@

\section{Participant information and data exclusion criteria}\label{a-participants}

This supplement provides information on the participants of the 12 experiments and the criteria by which participants' data were excluded. The columns of Table \ref{t:exclusion} show, for each of the 12 experiments (Exp.~1q, \ldots, Exp.~3c) how many participants were recruited, their age range and mean age, and their self-reported gender (`f' = female, `m' = male, `o' = other, `u' = undeclared). No gender data was collected in Exp.~1q. Participants' data were excluded based on the following criteria:

@@

\item `multiple': Due to an experimental glitch, some participants participated more than once in Exp.~1q. Since no information was available on which one was their first take, those participants' data was removed.

@@

\item `controls': Participants' data were excluded if their mean rating on the 6 main clause control items in the projection block was more than 2 sd above the group mean (in Exps.~1q, 2q, and 3q) or more than 2 sd below the group mean (in the remaining experiments). Participants' data were also excluded if their mean rating on the 6 main clause control items in the at-issueness block was more than 2 sd above the group mean (across all experiments).

\item `variance': Participants' data were excluded if they always selected roughly the same point on the response scale for the target stimuli, that is, if their mean variance on the target stimuli was more than 2 sd below the group mean variance. {\bf and manually inspecting their response patterns. Exps 1q, 2q, 3q: participants with low variance not excluded after manual inspection NEED TO CONSIDER THIS AGAIN}
\end{itemize}
